
# Individual-level ACTIONet analysis
# This script splits the integrated neuronal object by disease
# and then runs ACTIONet separately for each individual sample
# to obtain individual-level cluster representations.

# REneurons_ace is the integrated neuronal object before individual-level ACTIONet analysis.

#Running ACTIONet_MS
```{r}
unique(REneurons_ace$Disease)
REneurons_MS <- REneurons_ace[,REneurons_ace$Disease == "Ms_Neurons" ]
unique(REneurons_MS$Individual)
unique(colData(REneurons_MS)[,"Individual"])
colData(REneurons_ace)


MS_individual <- unique(colData(REneurons_MS)[,"Individual"])
MS_ACTIONet_run <- setNames(
      lapply(MS_individual, function(individual_i) {

       obj_i <- REneurons_MS[, colData(REneurons_MS)$Individual == individual_i, drop = FALSE]

         x <- ACTIONet::normalize.ace(obj_i)
         x <- ACTIONet::reduce.ace(x)
         x <- ACTIONet::runACTIONet(x, thread_no = 1)
         x <- ACTIONet::clusterCells(x)
   return(x)
    }
  ), 
MS_individual
 )
saveRDS(MS_ACTIONet_run, "path/to/MS_ACTIONet_run.rds")
MS_ACTIONet_run <- setNames(MS_ACTIONet_run,MS_individual)
#MS_ACTIONet_run <- readRDS('path/to/MS_ACTIONet_run.rds')

#MS
#MS_ACTIONet_run <- lapply(names(MS_ACTIONet_run), function(id){
    #obj <- MS_ACTIONet_run[[id]]
     # obj$leiden <- paste0(id, "_c", obj$leiden)
       #obj})

#MS_ACTIONet_run <- lapply(names(MS_ACTIONet_run), function(id){
   #obj <- MS_ACTIONet_run[[id]]
     #obj$leiden <- sub("^.*_c", "", obj$leiden)
      #obj})


```

#Running ACTIONet_ASD
```{r}

REneurons_autism <- REneurons_ace[,REneurons_ace$Disease == "autism_Neurons" ]
unique(REneurons_autism$Individual)
unique(colData(REneurons_autism)[,"Individual"])
colData(REneurons_ace)


ASD_individual <- unique(colData(REneurons_autism)[,"Individual"])
ASD_ACTIONet_run <- setNames(
      lapply(ASD_individual, function(individual_i) {

       obj_i <- REneurons_autism[, colData(REneurons_autism)$Individual == individual_i, drop = FALSE]

         x <- ACTIONet::normalize.ace(obj_i)
         x <- ACTIONet::reduce.ace(x)
         x <- ACTIONet::runACTIONet(x, thread_no = 1)
         x <- ACTIONet::clusterCells(x)
   return(x)
    }
  ), 
ASD_individual
 )
ASD_ACTIONet_run <- setNames(ASD_ACTIONet_run,ASD_individual)
#saveRDS(ASD_ACTIONet_run,'path/to/ASD_ACTIONet_run.rds')

#ASD_ACTIONet_run <- readRDS('path/to/ASD_ACTIONet_run.rds')


#ASD_ACTIONet_run <- lapply(names(ASD_ACTIONet_run), function(id){

#obj <- ASD_ACTIONet_run[[id]]
#obj$leiden <- sub("^.*_c", "", obj$leiden)
 #obj})

#ASD
#ASD_ACTIONet_run <- lapply(names(ASD_ACTIONet_run), function(id){

 # obj <- ASD_ACTIONet_run[[id]]
#obj$leiden <- paste0(id, "_c", obj$leiden)
# obj})
```



#Running ACTIONet_MDD
```{r}
unique(REneurons_ace$Disease)
REneurons_MDD <- REneurons_ace[,REneurons_ace$Disease == "MDD_Neurons" ]
unique(REneurons_MDD$Individual)
unique(colData(REneurons_MDD)[,"Individual"])
colData(REneurons_ace)


MDD_individual <- unique(colData(REneurons_MDD)[,"Individual"])
MDD_ACTIONet_run <- setNames(
      lapply(MDD_individual, function(individual_i) {

       obj_i <- REneurons_MDD[, colData(REneurons_MDD)$Individual == individual_i, drop = FALSE]

         x <- ACTIONet::normalize.ace(obj_i)
         x <- ACTIONet::reduce.ace(x)
         x <- ACTIONet::runACTIONet(x, thread_no = 1)
         x <- ACTIONet::clusterCells(x)
   return(x)
    }
  ), 
MDD_individual
 )
saveRDS(MDD_ACTIONet_run,'path/to/MDD_ACTIONet_run.rds')
#MDD_ACTIONet_run <- readRDS('path/to/MDD_ACTIONet_run.rds')
```


```{r}
#MDD
#MDD_ACTIONet_run <- lapply(names(MDD_ACTIONet_run), function(id){
 # obj <- MDD_ACTIONet_run[[id]]
  #obj$leiden <- paste0(id, "_c", obj$leiden)
 # obj})
#MDD_ACTIONet_run <- setNames(MDD_ACTIONet_run,MDD_individual)
```






#Running ACTIONet_AD
```{r}

unique(REneurons_ace$Disease)
REneurons_AD <- REneurons_ace[,REneurons_ace$Disease == "Alzheimer_Neurons" ]
unique(REneurons_AD$Individual)
unique(colData(REneurons_AD)[,"Individual"])
colData(REneurons_ace)


AD_individual <- unique(colData(REneurons_AD)[,"Individual"])
AD_ACTIONet_run <- setNames(
      lapply(AD_individual, function(individual_i) {

       obj_i <- REneurons_AD[, colData(REneurons_AD)$Individual == individual_i, drop = FALSE]

         x <- ACTIONet::normalize.ace(obj_i)
         x <- ACTIONet::reduce.ace(x)
         x <- ACTIONet::runACTIONet(x, thread_no = 1)
         x <- ACTIONet::clusterCells(x)
   return(x)
    }
  ), 
AD_individual
 )

AD_ACTIONet_run <- lapply(names(AD_ACTIONet_run), function(id){
  obj <- AD_ACTIONet_run[[id]]
  obj$leiden <- paste0(id, "_c", obj$leiden)
  obj 
})

#setNames
AD_ACTIONet_run <- setNames(AD_ACTIONet_run, AD_individual)

#saveRDS(AD_ACTIONet_run,'path/to/AD_ACTIONet_run.rds')
# AD_ACTIONet_run <- readRDS("path/to/AD_ACTIONet_run.rds") 

#AD
#AD_ACTIONet_run <- lapply(names(AD_ACTIONet_run), function(id){
   #obj <- AD_ACTIONet_run[[id]]
   # obj$leiden <- paste0(id, "_c", obj$leiden)
     #obj })





#AD
#AD_ACTIONet_run <- lapply(names(AD_ACTIONet_run), function(id){
  #obj <- AD_ACTIONet_run[[id]]
    #obj$leiden <- sub("^.*_c", "", obj$leiden)
      # obj})
```


