
# Pseudobulk construction and batch correction
# This script constructs pseudobulk profiles from cluster-level labels,
# assigns relevant metadata, and applies batch correction while preserving
# biological covariates such as cell type and diagnosis.

# Create the pseudobulks and normalize them

Functions to create the pseudobulk

```{r, basic functions}
# pairwise fold change
FC_pairwise <- function(expMat, QueryCol, RefColSet=NULL, sortedFC=F) {
  if(is.null(RefColSet)) RefColSet <- colnames(expMat)[-which(colnames(expMat)==QueryCol)]
  temp <- do.call(cbind, lapply(RefColSet, function(i) expMat[,QueryCol]-expMat[,i]))
  colnames(temp) <- RefColSet
  rownames(temp) <- rownames(expMat)
  FC <- rowSums(temp)
  if(sortedFC) FC <- sort(FC, decreasing = T)
  return(FC)
}

FC_signatures <- function(xM) {

  out <- do.call(
    cbind,
    lapply(base::colnames(xM), function(i) {
      FC_pairwise(xM, QueryCol = i, sortedFC = FALSE)
    })
  )

  base::colnames(out) <- base::colnames(xM)
  
  return(out)
}




# get counts for pseudobulk
get_column_group_count <- function(x, group, doParallel=FALSE) {
  Ugroups <- sort(names(which(table(group)>1)))
  if(doParallel) out <- do.call(cbind, parallel::mclapply(mc.cores = 8, X=Ugroups, function(i) Matrix::rowSums(as(x[,group==i], "sparseMatrix")))) else out <- do.call(cbind, lapply(Ugroups, function(i) Matrix::rowSums(as(x[,group==i], "sparseMatrix"))))
  colnames(out) <- Ugroups
  rownames(out) <- rownames(x)
  return(out)
}

# CPM normalisation
normalize_cpm <- function(M, logIt=T) {
  cpm <- t(t(M)/colSums(M, na.rm = T))*1000000
  if(logIt) cpm <- log2(cpm+1)
  return(cpm)
}

# pseudobulking and post hoc analysis
get_pseudobulk_sce <- function(xsce, xvec, ...) {
  # sum counts in each group given in xvec
  pseudobulk <- get_column_group_count(counts(xsce), xvec, ...)
  # normalize to get logcounts
  P <- normalize_cpm(pseudobulk)
  # get signatures of logcounts
  S <- FC_signatures(P)
  o <- SingleCellExperiment(assays=list(counts=pseudobulk, logcounts=P, signatures=S))
  if(is.null(colnames(o))) {
    colnames(o) <- colnames(pseudobulk)
  }
  o@metadata$cell_count <- table(xvec)
  return(o)
}
```

```{r, pseudobulk }
 z <- unname(REneurons_ace$Cluster_TAG)
pseudo_REneurons_sce <- get_pseudobulk_sce(REneurons_ace, z)

saveRDS(pseudo_REneurons_sce,'path/to/pseudo_REneurons_sce.rds')

```

##Adding ColData
```{r, colData}
#sample=Individual
#cluster=Cluster_TAG

pseudo_REneurons_sce$Individual <-  apply(table(REneurons_ace$Cluster_TAG, REneurons_ace$Individual)[colnames(pseudo_REneurons_sce),], 1, function(i) {colnames(table(REneurons_ace$Cluster_TAG, REneurons_ace$Individual))[which(i != 0)]})

                       
pseudo_REneurons_sce$CellType <- apply(table(REneurons_ace$Cluster_TAG, REneurons_ace$Cluster.Cell.Type)[colnames(pseudo_REneurons_sce),], 1, function(i) {colnames(table(REneurons_ace$Cluster_TAG, REneurons_ace$Cluster.Cell.Type))[which(i != 0)]})
unique(REneurons_ace$Cluster.Cell.Type)


pseudo_REneurons_sce$Study <- apply(table(REneurons_ace$Cluster_TAG, REneurons_ace$Study)[colnames(pseudo_REneurons_sce),], 1, function(i) {colnames(table(REneurons_ace$Cluster_TAG, REneurons_ace$Study))[which(i != 0)]})

```

##adding Diagnosis
```{r, colData}

Indivdual_Neuron_ace <- readRDS('path/to/Indivdual_Neuron_ACTIONetRUN_ace.rds')


Individual_Dictionary_Diagnosis<- setNames(Indivdual_Neuron_ace$Diagnosis, Indivdual_Neuron_ace$Individual)

pseudo_REneurons_sce$Diagnosis <- Individual_Dictionary_Diagnosis [pseudo_REneurons_sce$Individual]
unname(pseudo_REneurons_sce$Diagnosis )

#saveRDS(pseudo_REneurons_sce,'path/to/pseudo_REneurons_sce.rds')

```


#Batch Correction

##Diagnosis
```{r}
Diagnosis_batchcorrected_REneurons_sce <- ComBat_seq(counts = counts(pseudo_REneurons_sce), batch = pseudo_REneurons_sce$Study, covar_mod = data.frame(pseudo_REneurons_sce$CellType,pseudo_REneurons_sce$Diagnosis)) #works on counts! 

#saveRDS(Diagnosis_batchcorrected_REneurons_sce,'path/to/Diagnosis_batchcorrected_REneurons_sce.rds')


```
##adding the matrix to our psoudo-bulk
```{r}
#r, normalisation
pseudoCorrected_REneurons_sce <- pseudo_REneurons_sce
counts(pseudoCorrected_REneurons_sce) <- as(Diagnosis_batchcorrected_REneurons_sce, "sparseMatrix")
logcounts(pseudoCorrected_REneurons_sce) <- normalize_cpm(counts(pseudoCorrected_REneurons_sce))

saveRDS(pseudoCorrected_REneurons_sce,'path/to/pseudoCorrected_REneurons_sce.rds')
```
