
# Data preparation and integrated neuronal object construction
# This script loads multiple brain snRNA-seq datasets,
# extracts neuronal populations, identifies shared genes,
# and constructs an integrated ACTIONetExperiment object
# with relevant cell-level metadata.


# DataSet preparation for integrated neuronal object
##reading
```{r}

Alzheimer_ace <- readRDS("path/to/MathysDavila2019_AD.sce.rds")
MDD_ace <- readRDS("path/to/Nagy2020_MDD_PFC_snRNA.ace.rds")
MS_ace <- readRDS("path/to/Schirmer2019_MS.sce.rds")
autism_ace <- readRDS("path/to/Velmeshev2019_autism.sce.rds")
Jorstad_ace <- readRDS("path/to/Jorstad2023.PFC.sce.rds")
MDD_ace <- as.SingleCellExperiment(ace = MDD_ace)
```


##separating the Neurons
```{r}
#Alzheimer
unique(Alzheimer_ace$cluster.name)
Alzheimer_cells <- c("Ex7","Ex2","Ex1","Ex4","Ex8","Ex6","Ex3","Ex5","Ex10",
                      "In5","In3","In1","In2","In4" )
Alzheimer_Neurons <-  Alzheimer_ace[,Alzheimer_ace$cluster.name %in% Alzheimer_cells]

#MDD
unique(MDD_ace$reported.celltype)
MDD_Neurons <- MDD_ace[,MDD_ace$reported.celltype %in% c("Inhib","Ex")]

#MS
unique(MS_ace$cell_type)
Ms_cells <- c("EN-L5-6", "EN-L2-3-MS", "IN-PV", "EN-L4", "EN-MIX",
               "IN-SST", "IN-VIP", "EN-L2-3-Cntl", "IN-SV2C", "EN-PYR")
Ms_Neurons <- MS_ace[,MS_ace$cell_type %in% Ms_cells]

#autism
unique(autism_ace$cluster)
autism_cells <- c("Neu-NRGN-II", "L5/6", "IN-VIP", "L5/6-CC", 
                         "IN-SV2C", "L2/3", "IN-PV", "IN-SST", "Neu-mat", "Neu-NRGN-I", "L4")

autism_Neurons <-  autism_ace[ , autism_ace$cluster %in% autism_cells]
autism_Neurons$cluster <- droplevels(autism_Neurons$cluster)
table(autism_Neurons$cluster)

#Jorstad
unique(Jorstad_ace$cell_type)
Jorstad_cells <- c("Lamp5", "Sncg", "Vip", "Lamp5 Lhx6", "Pax6", "L5 ET", "L5/6 NP", 
             "L6 CT", "L6b", "L2/3 IT", "L4 IT", "L6 IT", "L5 IT", "L6 IT Car3", 
             "Chandelier", "Pvalb", "Sst", "Sst Chodl")
Jorstad_Neurons <- Jorstad_ace[,Jorstad_ace$cell_type %in% Jorstad_cells]

```

##find similar genes
```{r}
common_genes <- intersect(rownames(autism_Neurons),
                          intersect(rownames(Ms_Neurons), 
                                    intersect(rownames(MDD_Neurons),
                                              intersect(rownames(Alzheimer_Neurons), rownames(Jorstad_Neurons))
                                    )
                          )
)
common_genes
length(common_genes)
```


##make an ACTIONet experiment
```{r}
REneurons_ace <- ACTIONetExperiment::ACTIONetExperiment(
  assays = list(
    counts = cbind(
      counts(autism_Neurons)[common_genes, ],
      counts(Ms_Neurons)[common_genes, ],
      counts(MDD_Neurons)[common_genes, ],
      counts(Alzheimer_Neurons)[common_genes, ],
      counts(Jorstad_Neurons)[common_genes, ]
    )
  )
)
```

##TAG
```{r}
#ASD
ASDtag <- paste0(autism_Neurons$cell,"_ASD")
length(autism_Neurons$cell)

#MS
Ms_Neurons$cell
MStag <- paste0(Ms_Neurons$cell,"_MS")

#MDD
MDD_Neurons$barcode
MDDtag <- paste0(MDD_Neurons$barcode,"_MDD")

#AD
Alzheimer_Neurons$TAG
ADtag <- paste0(Alzheimer_Neurons$TAG,"_AD")

#JORSTAD
Jorstad_Neurons$cell_ID
JORStag <- paste0(Jorstad_Neurons$cell_ID,"_JORS")
length(JORStag)

CellTAGs <- c(ASDtag, MStag, MDDtag, ADtag, JORStag)
length(CellTAGs)
REneurons_ace$CellTAGs <- CellTAGs
```

##Diagnosis 
```{r}
#Alzheimer_Neurons
Alzheimer_Neurons$AD.group[Alzheimer_Neurons$AD.group %in% c("Mod-Severe-AD", "Pre-Mild-AD")] <- "AD"
Alzheimer_Neurons$AD.group[Alzheimer_Neurons$AD.group %in% "Non-AD"] <- "Control"

ADdiagnosis <- Alzheimer_Neurons$AD.group
class(ADdiagnosis)
table(ADdiagnosis)

#MDD_Neurons
table(MDD_Neurons$clinical.phenotype)
MDD_Neurons$clinical.phenotype[MDD_Neurons$clinical.phenotype %in% "Suicide"] <- "MDD"

MDDdiagnosis <- MDD_Neurons$clinical.phenotype
table(MDDdiagnosis)

#Ms_Neurons
table(Ms_Neurons$diagnosis)
MSdiagnosis <- Ms_Neurons$diagnosis

#autism_Neurons(ASD)
table(autism_Neurons$diagnosis)
ASDdiagnosis <- autism_Neurons$diagnosis
table(ASDdiagnosis)

#jorstad 
JORSdiagnosis <- rep("Control", length(Jorstad_Neurons$cell_type))

Diagnosis <- c(ASDdiagnosis, MSdiagnosis, MDDdiagnosis, ADdiagnosis, JORSdiagnosis)
length(Diagnosis)
REneurons_ace$Diagnosis <- Diagnosis
```

##Cell_type
```{r}
length(autism_Neurons$cluster)
autism_NeuronsCELLtype <- autism_Neurons$cluster
autism_neurons <- rep("autism_Neurons", length(autism_Neurons$cluster))

length(Ms_Neurons$cell_type)
Ms_neuronsCELLtype <- Ms_Neurons$cell_type
Ms_neurons <- rep("Ms_Neurons", length(Ms_Neurons$cell_type))

length(MDD_Neurons$reported.celltype)
MDD_neuronsCELLtype <- MDD_Neurons$reported.celltype
MDD_neurons <- rep("MDD_Neurons", length(MDD_Neurons$reported.celltype))

length(Alzheimer_Neurons$cluster.name)
Alzheimer_neuronsCELLtype <- Alzheimer_Neurons$cluster.name
Alzheimer_neurons <- rep("Alzheimer_Neurons", length(Alzheimer_Neurons$cluster.name))

length(Jorstad_Neurons$cell_type)
Jorstad_NeuronCelltype <- Jorstad_Neurons$cell_type
Jorstad_neurons <- rep("Jorstad_Neurons", length(Jorstad_Neurons$cell_type))
```

## Metadata assignment
```{r}
REneurons_ace$Disease <- c(autism_neurons, Ms_neurons, MDD_neurons, Alzheimer_neurons, Jorstad_neurons)

REneurons_ace$Cell.Type <- c(
  as.vector(autism_NeuronsCELLtype),
  as.vector(Ms_neuronsCELLtype),
  as.vector(MDD_neuronsCELLtype),
  as.vector(Alzheimer_neuronsCELLtype),
  as.vector(Jorstad_NeuronCelltype)
)

REneurons_ace$Diagnosis <- c(
  as.vector(ASDdiagnosis),
  as.vector(MSdiagnosis),
  as.vector(MDDdiagnosis),
  as.vector(ADdiagnosis),
  as.vector(JORSdiagnosis)
)

REneurons_ace$CellTAGs <- c(ASDtag, MStag, MDDtag, ADtag, JORStag)

Jorstad_NA <- rep("NA", length(REneurons_ace$Cell.Type) - length(Jorstad_NeuronCelltype))
REneurons_ace$Jorstad <- c(as.vector(Jorstad_NA), as.vector(Jorstad_NeuronCelltype))
```

##Individual 
```{r}
#autism individual
colData(autism_Neurons)
table(autism_Neurons$individual, autism_Neurons$age)

unique(REneurons_ace$Disease)
autism_TAGs <- REneurons_ace$CellTAGs[REneurons_ace$Disease == "autism_Neurons"]
unique(REneurons_ace$CellTAGs)
dim(autism_Neurons)
length(autism_TAGs)
orig_autism <- sub("_ASD", "", autism_TAGs)
unique(orig_autism == autism_Neurons$cell)

autism_individual <- autism_Neurons$individual[match(orig_autism, autism_Neurons$cell)]
length(autism_individual)
length(autism_Neurons$cell)
length(autism_TAGs)
autism_individual <- paste0("ASD_", autism_individual)

#MS individual
colData(Ms_Neurons)
table(Ms_Neurons$sample, Ms_Neurons$age)

unique(REneurons_ace$Disease)
MS_TAGs <- REneurons_ace$CellTAGs[REneurons_ace$Disease == "Ms_Neurons"]
dim(Ms_Neurons)
length(MS_TAGs)
orig_MS <- sub("_MS(?!.*_MS)", "", MS_TAGs, perl = TRUE)
length(orig_MS)
length(Ms_Neurons$cell)
unique(orig_MS == Ms_Neurons$cell)
unique(orig_MS %in% Ms_Neurons$cell)
unique(Ms_Neurons$sample)
colData(Ms_Neurons)

MS_individual <- Ms_Neurons$sample[match(orig_MS, Ms_Neurons$cell)]
length(autism_individual)
length(autism_TAGs)
MS_individual <- paste0("MS_", MS_individual)

#MDD individual
colData(MDD_Neurons)
table(MDD_Neurons$subject.id, MDD_Neurons$batch)

unique(REneurons_ace$Disease)
MDD_TAGs <- REneurons_ace$CellTAGs[REneurons_ace$Disease == "MDD_Neurons"]
orig_MDD <- sub("_MDD", "", MDD_TAGs)
dim(MDD_Neurons)
length(MDD_TAGs)
length(orig_MDD)

MDD_individual <- MDD_Neurons$subject.id[match(orig_MDD, MDD_Neurons$barcode)]
length(MDD_individual)
length(orig_MDD)
MDD_individual <- paste0("MDD_", MDD_individual)

#Alzheimer individual
unique(REneurons_ace$Disease)
Alzheimer_TAGs <- REneurons_ace$CellTAGs[REneurons_ace$Disease == "Alzheimer_Neurons"]
orig_Alzheimer <- sub("_AD", "", Alzheimer_TAGs)
length(Alzheimer_TAGs)
length(orig_Alzheimer)
unique(orig_Alzheimer == Alzheimer_Neurons$TAG)

Alzheimer_individual <- Alzheimer_Neurons$age_bl[match(orig_Alzheimer, Alzheimer_Neurons$TAG)]
length(Alzheimer_individual)
length(orig_Alzheimer)
Alzheimer_individual <- paste0("AD_", Alzheimer_individual)

#Jorstad individual
unique(REneurons_ace$Disease)
Jorstad_TAGs <- REneurons_ace$CellTAGs[REneurons_ace$Disease == "Jorstad_Neurons"]
orig_Jorstad <- sub("_JORS", "", Jorstad_TAGs)
length(Alzheimer_TAGs)
length(orig_Alzheimer)
unique(orig_Jorstad == Jorstad_Neurons$cell_ID)
colData(Jorstad_Neurons)

Jorstad_individual <- Jorstad_Neurons$batch[match(orig_Jorstad, Jorstad_Neurons$cell_ID)]
length(Jorstad_individual)
length(orig_Jorstad)
Jorstad_individual <- paste0("Jors_", Jorstad_individual)

#make one vector
Individual <- c(autism_individual, MS_individual, MDD_individual, Alzheimer_individual, Jorstad_individual)
saveRDS(Individual, "outputs/Individual.RDS")

unique(Individual)
length(Individual)
length(REneurons_ace$Cell.Type)
REneurons_ace$Individual <- Individual
```


##Double_Check
```{r}
table(REneurons_ace$Disease)

#ASD
table(REneurons_ace$Cell.Type[which(REneurons_ace$Disease == "autism_Neurons")])
table(autism_Neurons$cluster)
table(autism_Neurons$cluster) == table(REneurons_ace$Cell.Type[which(REneurons_ace$Disease == "autism_Neurons")])

#MS
table(REneurons_ace$Cell.Type[which(REneurons_ace$Disease == "Ms_Neurons")])
table(Ms_Neurons$cell_type)
table(Ms_Neurons$cell_type) == table(REneurons_ace$Cell.Type[which(REneurons_ace$Disease == "Ms_Neurons")])

#MDD
table(REneurons_ace$Cell.Type[which(REneurons_ace$Disease == "MDD_Neurons")])
table(MDD_Neurons$reported.celltype)
table(MDD_Neurons$reported.celltype) == table(REneurons_ace$Cell.Type[which(REneurons_ace$Disease == "MDD_Neurons")])

#AD
table(REneurons_ace$Cell.Type[which(REneurons_ace$Disease == "Alzheimer_Neurons")])
table(Alzheimer_Neurons$cluster.name)
table(Alzheimer_Neurons$cluster.name) == table(REneurons_ace$Cell.Type[which(REneurons_ace$Disease == "Alzheimer_Neurons")])

#JORS
table(REneurons_ace$Cell.Type[which(REneurons_ace$Disease == "Jorstad_Neurons")])
table(Jorstad_Neurons$cell_type)
table(as.vector(Jorstad_Neurons$cell_type)) == table(REneurons_ace$Cell.Type[which(REneurons_ace$Disease == "Jorstad_Neurons")])
```


